A Logical Solution of the Linearity Challenge

### 2025-07-21 paper/calculus.tex
progress

The calculus of pure sessions that is formalized in our development is dubbed
\Calculus and is a linear version of \CP~\citep{Wadler14,GayVasconcelos25}. Its
syntax makes use of an infinite set of \emph{channel names}, ranged over by $x$,
$y$ and $z$, and is shown in Table~\ref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the $x$ and $y$ channels, so that
every message sent on one of the channels is forwarded to the other. This form
is useful for modeling \emph{free outputs}, as we will see later.\Luca{Mettere
esempio!}
%
The $\Close\x$ and $\Wait\x.P$ processes respectively model the output and input
of a termination signal on the channel $x$. The former process has no
continuation, while the latter process continues as $P$ once the signa has been
received.
%
The $\Fail\x$ process denotes a runtime error concerning the channel $x$.
%
The $\Select\x{\InTag_i}\z.P$ and $\Case\x\z{Q_1}{Q_2}$ processes respectively
denote the output and input of a label $\InTag_i$ along with a fresh
continuation $z$ on the channel $x$. Depending on the label $\InTag_i$, the
receiver continues as $Q_i$. In practice it is useful to consider arbitrary
(domain-specific) sets of tags instead of the canonical two $\InTag_1$ and
$\InTag_2$. This generalization has no substantial impact on the formalization,
so we stick to the classical case of ``binary choices'' in line with the usual
presentations of linear logic.
%
The $\Fork\x\y\z{P}{Q}$ and $\Join\x\y\z.R$ processes respectively denote the
output and input of a pair of fresh channels $y$ and $z$ on the channel $x$. The
former process forks into $P$ and $Q$, each using $y$ and $z$ respectively. The
latter process continues as $R$. These forms are often used in conjunction with
links to model \emph{free outputs}, but can also be used to model bifurcating
sessions whereby a sequential protocol continues as two parallel sub-protocols.

@@

\emph{Cuts} of the form $\Cut\x{P}{Q}$ represent the parallel composition of $P$
and $Q$ connected by a channel $x$.

The notions of free and bound names are fairly standard. Just bear in mind that
all action prefixes bind continuation channels in (some) continuation processes.
In particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, $\Case\x\z{P}{Q}$ binds
$z$ in both $P$ and $Q$, $\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and
$z$ in $Q$ but not in $P$, and $\Join\x\y\z.P$ binds both $y$ and $z$ in $P$.
%
Also, $\Server\x\y.P$ and $\Client\x\y.P$ bind $y$ in $P$ and
$\Contract\x\y\z.P$ binds $y$ and $z$ in $P$.
%
We write $\fn{P}$ for the set of channel names occurring free in $P$ and we
identify processes up to renaming of bound names.

@@

The operational semantics of \Calculus is shown in Table~\ref{tab:semantics} and
is given by two relations: a \emph{structural pre-congruence} relation $\pcong$
and a \emph{reduction} relation $\red$. Structural pre-congruence relates
processes that we consider to be essentially indistinguishable. Reduction
relates processes in which a communication occurs. In some presentations of
linear logic~\cite{Doumane}, these relations are respectively dubbed
\emph{external} and \emph{internal} reductions. Let us describe the two
relations more in detail.

Structural pre-congruence is the least pre-congruence defined by rule
\rulename{s-*} rules.
%
\SLink and \SComm assert that links and parallel compositions are symmetric. The
remaining rules, when read from left to right, push a cut on $x$ underneath the
topmost prefix on $y$ of one of its sub-processes when $x \ne y$. These rules
are key to float input/output actions to the top-level of a process, so that
they can interact with corresponding complementary actions in the context
surrounding the process.
%
Note that some rules have additional side-conditions preventing bound names to
become free.\Luca{Non proprio, sono tutti nomi bound.} Also, there are two
versions of \SForkL and \SForkR depending on which of the two continuations $P$
and $Q$ contains occurrences of the restricted channel $x$.
%
Notably and somewhat surprisingly, there is no pre-congruence rule asserting
that parallel composition is associative. While this is a sound and expected
property of parallel composition, it turns out to be inessential for proving all
of the properties of \Calculus that we consider. We leave it out in light of the
forthcoming formalization of the calculus, in which every simplification is
welcome, although standard presentations of classical linear logic admit such
reduction~\cite{Doumane17}.

The reduction rules \RLink, \RClose, \RSelect and \RFork are in one-to-one
correspondence with the well-known reductions of session calculi and the cut
reductions of linear logic. \RLink eliminates a link $\Link\x\y$ by unifying $x$
and $y$. \RClose closes a session $x$ when a termination signal is exchanged on
$x$. \RSelect and \RFork respectively model the communication of a tag $\Tag_i$
and of a fresh channel $y$. Since channels are linear in \Calculus, each
non-terminating communication carries along a \emph{fresh continuation} $z$ on
which the rest of the conversation may take place, in line with the semantics of
the linear $\pi$-calculus~\cite{KobayashiPierceTurner99}.
%
The rule~\RCut propagates reductions through cuts and the rules \RCong closes
reduction by structural pre-congruence.

\subsection{Type System}

The types for \Calculus, ranged over by $A$, $B$, \dots, are propositions of
\MALL generated by the following grammar:
\[
    A, B ::= \Zero \mid \Top \mid \One \mid \Bot \mid A \Plus B \mid A \With B \mid A \Ten B \mid A \Par B
\]

The interpretation of propositions as session types is the usual
one~\cite{Wadler14}: the constants $\Bot$ and $\One$ describe the behavior of a
process sending/receiving a termination signal; the additive connectives $A
\Plus B$ and $\A \With B$ describe the behavior of a process sending/receiving a
boolean value and then adhering to either $A$ and $B$ accordingly; the
multiplicative connectives $A \Ten B$ and $A \Par B$ describe the behavior of a
process sending/receiving a channel (to be used according to $A$) and then
continuing as described by $B$. The additive constants $\Zero$ and $\Top$ do not
describe useful protocols, but they can serve the role of smallest/largest type
in type systems that support a notion of subtyping~\citep{HornePadovani24}.

@@

context syntax and notation

>>>>

The calculus of pure sessions that is formalized in our development is dubbed
\Calculus and is a linear version of \CP~\citep{Wadler14,GayVasconcelos25}. Its
syntax makes use of an infinite set of \emph{channel names}, ranged over by $x$,
$y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the $x$ and $y$ channels, so that
every message sent on one of the channels is forwarded to the other. This form
is useful for modeling \emph{free outputs}, as we will see later.\Luca{Mettere
esempio!}
%
The $\Close\x$ and $\Wait\x.P$ processes respectively model the output and input
of a termination signal on the channel $x$. The former process has no
continuation, while the latter process continues as $P$ once the signa has been
received.
%
The $\Fail\x$ process denotes a runtime error concerning the channel $x$.
%
The $\Select\x{\InTag_i}\z.P$ and $\Case\x\z{Q_1}{Q_2}$ processes respectively
denote the output and input of a label $\InTag_i$ along with a fresh
continuation $z$ on the channel $x$. Depending on the label $\InTag_i$, the
receiver continues as $Q_i$. In practice it is useful to consider arbitrary
(domain-specific) sets of tags instead of the canonical two $\InTag_1$ and
$\InTag_2$. This generalization has no substantial impact on the formalization,
so we stick to the classical case of ``binary choices'' in line with the usual
presentations of linear logic.
%
The $\Fork\x\y\z{P}{Q}$ and $\Join\x\y\z.R$ processes respectively denote the
output and input of a pair of fresh channels $y$ and $z$ on the channel $x$. The
former process forks into $P$ and $Q$, each using $y$ and $z$ respectively. The
latter process continues as $R$. These forms are often used in conjunction with
links to model \emph{free outputs}, but can also be used to model bifurcating
sessions whereby a sequential protocol continues as two parallel sub-protocols.

@@

\emph{Cuts} of the form $\Cut\x{P}{Q}$ represent the parallel composition of the
processes $P$ and $Q$ connected by a channel $x$.

The notions of free and bound names are almost standard. Just bear in mind that
also output action prefixes bind continuation channels in (some) continuation
processes. In particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, while
$\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in
$P$.
%
We write $\fn{P}$ for the set of channel names occurring free in $P$ and we
identify processes up to renaming of bound names.

@@

The operational semantics of \Calculus is shown in~\Cref{tab:semantics} and
is given by two relations: a \emph{structural pre-congruence} relation $\pcong$
and a \emph{reduction} relation $\red$. Structural pre-congruence relates
processes that we consider to be essentially indistinguishable. Reduction
relates processes in which a communication occurs. In some presentations of
linear logic~\cite{BaeldeDoumaneSaurin16,Doumane17}, these relations are
respectively dubbed \emph{external} and \emph{internal} reductions. Let us
describe the two relations more in detail.

Structural pre-congruence is the least pre-congruence defined by the
\rulename{s-*} rules.
%
\SLink and \SComm assert that links and parallel compositions are symmetric. The
remaining rules, when read from left to right, push a cut on $x$ underneath the
topmost prefix on $y$ of one of its sub-processes when $x \ne y$. These rules
are key to float input/output actions to the top-level of a process, so that
they can interact with corresponding complementary actions in the context
surrounding the process.
%
Note that some rules have additional side-conditions aimed at preserving the
meaning of names when binders are moved around. Also, there are two versions of
\SForkL and \SForkR depending on which of the two continuations $P$ and $Q$
contains occurrences of the restricted channel $x$.

Notably and somewhat surprisingly, there is no pre-congruence rule asserting
that parallel composition is associative. While this is a sound and expected
property of parallel composition, it turns out to be inessential for proving all
of the properties of \Calculus that we consider. We leave it out in light of the
forthcoming formalization of the calculus, in which every simplification is
welcome, although standard presentations of classical linear logic admit such
reduction~\cite{Doumane17}.
%
Another rule that deserves attention is \SServer. In this case, a server prefix
$\Server\y\u$ can be pulled out of a cut only provided that the other process in
the cut is also a server on the channel restricted by the cut.

The reduction rules \RLink, \RClose, \RSelect, \RFork, \RConnect, \RWeaken and
\RContract are in one-to-one correspondence with the well-known reductions of
session calculi and the principal cut reductions of linear logic. \RLink
eliminates a link $\Link\x\y$ by substituting $y$ for $x$. \RClose closes a
session $x$ when a termination signal is exchanged on $x$. \RSelect and \RFork
respectively model the communication of a tag $\Tag_i$ and of a fresh channel
$y$. Each of these communications carries along a \emph{fresh continuation} $z$
on which the rest of the conversation takes place, in line with the semantics of
the linear $\pi$-calculus~\cite{KobayashiPierceTurner99}.
%
\RConnect, \RWeaken and \RContract describe interactions on shared channels and
respectively represent the connection between a server and a client, the
disposal of an unused server, and the duplication of a server. In \RWeaken and
\RContract we use some convenient (albeit slightly ambiguous) notation for
denoting sequences of (pairwise distinct) channels and prefixes. In particular,
$\seqof\z$ stands for a sequence $z_1, \dots, z_n$ of channels,
$\Weaken{\seqof\z}.Q$ stands for $\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for
$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$.

The rule~\RCut propagates reductions through cuts and the rules \RCong closes
reduction by structural pre-congruence.

\subsection{Type System}
\label{sec:typing-rules}

The types for \Calculus, ranged over by $A$, $B$, \dots, are linear logic
 propositions generated by the following grammar:
\[
    A, B ::= \Zero \mid \Top \mid \One \mid \Bot \mid A \Plus B \mid A \With B \mid A \Ten B \mid A \Par B \mid \OfCourse A \mid \WhyNot A
\]

The interpretation of propositions as session types is the usual
one~\cite{Wadler14}: the constants $\Bot$ and $\One$ describe the behavior of a
process sending/receiving a termination signal; the additive connectives $A
\Plus B$ and $\A \With B$ describe the behavior of a process sending/receiving a
boolean value and then adhering to either $A$ and $B$ accordingly; the
multiplicative connectives $A \Ten B$ and $A \Par B$ describe the behavior of a
process sending/receiving a channel (to be used according to $A$) and then
continuing as described by $B$. The additive constants $\Zero$ and $\Top$ do not
describe useful protocols, but they can serve the role of smallest/largest type
in type systems that support a notion of subtyping~\citep{HornePadovani24}. The
``of course'' modality $\OfCourse A$ and the ``why not'' modality $\WhyNot A$
describe the behavior of servers and clients accepting and requesting
connections to be used according to $A$.

@@

Typing contexts (\ie sequents) are finite maps from channel names to types
written as $x_1 : A_1, \dots, x_n : A_n$ and ranged over by $\ContextC$ and
$\ContextD$. We write $\ContextC, \ContextD$ for the union of $\ContextC$ and
$\ContextD$ when they have disjoint domain. We write $\WhyNot\Context$ for some
context $\Context = x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the
types in its range are prefixed by the $\WhyNot$ modality.

\begin{table}
    \caption{Typing rules of \Calculus.}
    \label{tab:typing-rules}
    \begin{tabular}{@{}c@{}}
        \begin{mathpar}
            \inferrule[\LinkRule]{~}{
                \wtp{\Link\x\y}{x : A, y : \dual{A}}
            }
            \and
            \inferrule[\FailRule]{~}{
                \wtp{\Fail\x}{x : \Top, \Context}
            }
            \and
            \inferrule[\CloseRule]{~}{
                \wtp{\Close\x}{x : \One}
            }
            \and
            \inferrule[\WaitRule]{
                \wtp{P}{\Context}
            }{
                \wtp{\Wait\x.P}{x : \Bot, \Context}
            }
            \and
            \inferrule[\SelectRule]{
                \wtp{P}{y : A_i, \Context}
            }{
                \wtp{\Select\x{\InTag_i}\y.P}{x : A_1 \Plus A_2, \Context}
            }
            \and
            \inferrule[\CaseRule]{
                \wtp{P}{y : A, \Context}
                \\
                \wtp{Q}{y : B, \Context}
            }{
                \wtp{\Case\x\y{P}{Q}}{x : A \With B, \Context}
            }
            \and
            \inferrule[\ForkRule]{
                \wtp{P}{y : A, \ContextC}
                \\
                \wtp{Q}{z : B, \ContextD}
            }{
                \wtp{\Fork\x\y\z{P}{Q}}{x : A \Ten B, \ContextC, \ContextD}
            }
            \and
            \inferrule[\JoinRule]{
                \wtp{P}{y : A, z : B, \Context}
            }{
                \wtp{\Join\x\y\z.P}{x : A \Par B, \Context}
            }
            \and
            \inferrule[\ServerRule]{
                \wtp{P}{y : A, \WhyNot\Context}
            }{
                \wtp{\Server\x\y.P}{x : \OfCourse A, \WhyNot\Context}
            }
            \and
            \inferrule[\ClientRule]{
                \wtp{P}{y : A, \Context}
            }{
                \wtp{\Client\x\y.P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\WeakenRule]{
                \wtp{P}\Context
            }{
                \wtp{P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\ContractRule]{
                \wtp{P}{y : \WhyNot A, z : \WhyNot A, \Context}
            }{
                \wtp{P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\CutRule]{
                \wtp{P}{x : A, \ContextC}
                \\
                \wtp{Q}{x : \dual{A}, \ContextD}
            }{
                \wtp{\Cut\x{P}{Q}}{\ContextC, \ContextD}
            }
        \end{mathpar}
    \end{tabular}
\end{table}

Typing judgments have the form $\wtp{P}\Context$ meaning that the process $P$ is
well typed in the context $\Context$. Equivalently, the judgment indicates that
the sequent $\wtp{}\Context$ is derivable and $P$ is a \emph{proof term}
corresponding to the derivation for $\wtp{}\Context$.
%
The typing rules are shown in \Cref{tab:typing-rules}. Since they are in
one-to-one correspondence with the proof rules of classical linear logic, the
reader may refer to the standard literature on linear logic~\cite{} or
propositions as sessions~\citep{Wadler14} for their interpretation.

\subsection{Properties of Well-Typed Processes}
\label{sec:properties}

\begin{lemma}
    \label{lem:pcong}
    If $P \pcong Q$, then $\wtp{P}\Context$ implies $\wtp{Q}\Context$.
\end{lemma}

\begin{theorem}[subject reduction]
    \label{thm:red}
    If $P \red Q$, then $\wtp{P}\Context$ implies $\wtp{Q}\Context$.
\end{theorem}

\begin{theorem}[deadlock freedom]
    \label{thm:df}
    If $\wtp{P}\Context$, then $P$ is either observable or reducible.
\end{theorem}

safety
